Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-12-16 ms.tex
Ia

The source term in the continuity equation (eq.~[\ref{eq:drhodt}]),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$
years.  To good approximation $\eta\simeq 0.02 (t/t_h)^{-1.3}$ at time $t$ following a burst of star formation (e.g., \citet{Ciotti+91}),\footnote{Note that 
\citet{Ciotti+91} gives the overall mass return rate from evolved
stars as a function of B-band luminosity. We give the mass return rate
per unit volume. However, our expressions are equivalent.} although $\eta$ will generally be higher in cases of ongoing star formation.

@@

Energy injected by Type Ia supernovae (SNe Ia) represents a
potentially important source of heating, which unlike core collapse
SNe is present even in an evolved stellar population.  Assuming that
each Ia injects a thermal energy $E_{\rm Ia}$ into the interstellar
medium, and that the SNe rate (per stellar mass) is given by $R_{\rm
  Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm
  Ia}$ produces an effective wind heating parameter
(equation~[\ref{eq:vw_eff}])
\begin{align} v_{\rm w,0}^{\rm Ia} =\sqrt{\frac{2 t_h R_{\rm Ia}
E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align} The thermal energy injected by a Ia SN $E_{\rm Ia} \simeq
\epsilon_{\rm Ia} 10^{51}$ ergs depends on the efficiency $\epsilon$
with which the blast wave energy is converted into bulk or turbulent
motion instead of being lost to radiation.  \cite{Thornton+98}
estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$,
depending weakly on surrounding density, but \citet{Sharma+14} argues
that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if
the SNe occur in a hot dilute medium, as may characterize the circumnuclear medium.

@@

  Using equations (\ref{eq:vw_sne}), (\ref{eq:DTD}), and
  (\ref{eq:eta}) we thus estimate that

  \begin{align} 
    v_{w,0}^{\rm Ia} \approx 700(\epsilon_{\rm
      Ia}/0.4)^{0.5}(\tau_{\star}/t_{\rm h})^{0.09} \eta_{0.02}^{-1/2}\,{\rm km
      \,s^{-1}},
  \end{align}

 showing that the effective
  velocity due to SN heating is (coincidentally) a weak function of
  the stellar age.

  The relatively high value of $v_{w,0}^{\rm Ia}$ implies that Ia SNe
  represent a potentially dominant source of non-AGN heating in the
  CNM.  However, SNe can only be approximated as "constant" heating
  source throughout space and time if the rate of SNe is rapid
  compared to the inflow time at the radius of interest
  (\citealt{ShcherbakovWong+:2014a}).  To estimate when this
  approximation is valid, we estimate the ``Ia radius"
  \begin{align}
    r_{\rm Ia} \sim \left(\frac{G}{R_{\rm Ia}\sigma}\right)^{1/2} \sim
    38 M_{\bullet,8}^{-0.1}(\tau_{\star}/t_{\rm h})^{0.56}\,{\rm pc}
    \label{eq:rIa}
  \end{align}
as the location exterior of which the time interval between subsequent
supernovae $\tau_{\rm Ia} \sim 1/M_{\rm enc}R_{\rm Ia} \sim
G/r\sigma^{2}R_{\rm Ia}$ exceeds the local dynamical timescale $t_{\rm
dyn} \sim r/\sigma$, where we again adopt the Ia rate for an old
stellar population and in the final equality we estimate the velocity
dispersion using the $M_{\bullet}-\sigma$ relationship
(equation~[\ref{eq:Msigma}]).  Between subsequent supernova, stars
release a gaseous mass $M_{\rm gas} \approx \eta M_{\star}\tau_{\rm
Ia}/t_{\rm h}$ interior to the Ia radius, which is gravitationally
bound to the SMBH by an energy $E_{\rm bind} \approx M_{\rm
gas}\sigma^{2}$.  Thus, according to the above definitions, $E_{\rm
Ia}/E_{\rm bind} \sim (v_{\rm w,0}^{\rm Ia})^{2}/2\sigma^{2}$.

Note that the stagnation radius scales approximately linearly with
$M_{\bullet}$ (equation~[\ref{eq:stag_simple}]), while $\rIa$ is
roughly indpendent of SMBH mass.  For small $\Mbh$, the stagnation
radius thus lies well inside $\rIa$ and hence Ia SNe will not
contribute appreciably to the heating rate.  On the other hand, for
massive SMBHs $M_{\bullet} \gtrsim 10^8 \Msun$, SNe Ia represent an
important, and possibly dominant, contribution to the overall heating.

>>>>

The source term in the continuity equation (eq.~[\ref{eq:drhodt}]),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$
years.  To good approximation $\eta\simeq 0.02 (t_{\star}/t_h)^{-1.3}$ at time $t_{\star}$ following a burst of star formation (e.g., \citet{Ciotti+91}),\footnote{Note that 
\citet{Ciotti+91} gives the overall mass return rate from evolved
stars as a function of B-band luminosity. We give the mass return rate
per unit volume. However, our expressions are equivalent.} although $\eta$ will generally be higher in cases of ongoing star formation.

@@

Energy injected by Type Ia supernovae (SNe Ia) represents a potentially important source of heating, which unlike core collapse SNe is present even in an evolved stellar population.  Assuming that each Ia injects a thermal energy $E_{\rm Ia}$ into the interstellar medium, and that the SNe rate (per stellar mass) is given by $R_{\rm   Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm  Ia}$ produces an effective wind heating parameter (equation~[\ref{eq:vw_eff}]) \begin{align} v_{\rm w,0}^{\rm Ia} =\sqrt{\frac{2 t_h R_{\rm Ia}
E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align} The thermal energy injected by a Ia SN $E_{\rm Ia} \simeq
\epsilon_{\rm Ia} 10^{51}$ ergs depends on the efficiency $\epsilon$
with which the initial blast wave energy is converted into bulk or turbulent
motion instead of being lost to radiation.  \cite{Thornton+98}
estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$,
depending weakly on surrounding density, but \citet{Sharma+14} argues
that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if
the SNe occur in a hot dilute medium, as may characterize the circumnuclear medium.

@@

From equations (\ref{eq:vw_sne}), (\ref{eq:DTD}) we thus estimate that 
  \begin{eqnarray} 
    v_{w,0}^{\rm Ia} &\approx& 700(\epsilon_{\rm
      Ia}/0.4)^{0.5}(\tau_{\star}/t_{\rm h})^{-0.56}\eta_{0.02}^{-1/2}\,{\rm km
      \,s^{-1}} \nonumber \\
&\approx& 700(\epsilon_{\rm
      Ia}/0.4)^{0.5}(\tau_{\star}/t_{\rm h})^{0.09}\,{\rm km
      \,s^{-1}},
\label{eq:vIa}
  \end{eqnarray}
where in the second line we have taken $\eta\simeq 0.02 (\tau_{\star}/t_h)^{-1.3}$ for a single burst of star formation (\citealt{Ciotti+91}).

The high value of $v_{w,0}^{\rm Ia}$ implies that Ia SNe represent a potentially dominant source of non-AGN heating in the CNM.  However, SNe can only be approximated as being a constant heating source throughout space and time if the rate of SNe is rapid compared to the inflow time at the radius of interest (\citealt{ShcherbakovWong+:2014a}).  To estimate when this approximation is valid, we estimate the ``Ia radius"
  \begin{align}
    r_{\rm Ia} \sim \left(\frac{G}{R_{\rm Ia}\sigma}\right)^{1/2} \sim
    38 M_{\bullet,8}^{-0.1}(\tau_{\star}/t_{\rm h})^{0.56}\,{\rm pc}
    \label{eq:rIa}
  \end{align}
as the location exterior of which the time interval between subsequent supernovae $\tau_{\rm Ia} \sim 1/M_{\rm enc}R_{\rm Ia} \sim G/r\sigma^{2}R_{\rm Ia}$ exceeds the local dynamical timescale $t_{\rm
dyn} \sim r/\sigma$, where we again adopt the Ia rate for an old stellar population and in the final equality we estimate the velocity dispersion using the $M_{\bullet}-\sigma$ relationship
(equation~[\ref{eq:Msigma}]).  Between subsequent supernovae, stars release a gaseous mass $M_{\rm gas} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$ interior to the Ia radius, which is gravitationally bound to the SMBH by an energy $E_{\rm bind} \approx M_{\rm gas}\sigma^{2}$.  Thus, according to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{\rm w,0}^{\rm Ia})^{2}/2\sigma^{2}$.

Assuming that $v_{\rm w,0}^{\rm Ia} \gg \sigma$ then by substituting equation (\ref{eq:vIa}) into (\ref{eq:rs_simple}) we can write
\begin{eqnarray}
\frac{r_{\rm Ia}}{r_{\rm s}} &\approx& 12.4M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4)^{-1}(\tau_{\star}/t_{\rm h})^{1.68}\eta_{0.02}  \nonumber \\
&\approx& 12.4M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4)^{-1}(\tau_{\star}/t_{\rm h})^{0.38}
\end{eqnarray}
Thus only for massive SMBHs $M_{\bullet} \gtrsim 10^8 \Msun$ and/or very young populations will SNe Ia represent a steady source of heating at the nominal stagnation radius.
